---
title: "Intro to Data Frames in R"
author: "Environmental Data Analytics | John Fay and Luana Lima"
date: "1/10/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data frame definition

A **data frame** is used for storing data tables. It a list a vector of equal length. 

Special property: more general structures. You can have a combination of different modes.

When we import data to R, data frame is the preferred way for storing the data because columns can have different modes (character, numeric, integer, logical, complex).

ONE RESTRICTION - all of the columns need to have the same number of rows.

## Data frame built-in example

Let's look into a built-in data frame from package "datasets" - cars. The data give the speed of cars and the distances taken to stop.

* code chunks can be added using the green button. You can use the drop down menu to add different types of code, but you must have that code language installed to run it.

(cars is a pre-installed dataset in R)

```{r cars}
cars
```

Note that it has 2 columns and 50 rows.

We can tell that this is a data frame by its layout
dbl - means that it is a real number (can have decimal points)

## Data frame columns

Suppose you want just the column speed. How would you access that data? 

```{r}
cars$speed
```

*R will auto-populate the data sets that are available within that dataframe

How would you store it on another object?

```{r}
car_speed <- cars$speed
car_speed
```

<- is the assign symbol
command + return runs the single line

## Transforming object in a data frame

Is the new object you create a data frame?
The following commands are 2 ways you can check whether or not the object is a dataframe

```{r}
#Option 1
class(car_speed)

#Option 2
is.data.frame(car_speed)
```

How could you make it a data frame?
1. create new object using <-
2. call a function using the function name
3. put the object 

```{r}
df_car_speed <- as.data.frame(car_speed)
df_car_speed
class(df_car_speed)
```

object name was assigned as the column name within the new dataframe

## Adding columns to a data frame

How could you add columns to *df_car_speed*?

```{r}
car_dist <- cars$dist #pulling distance data from cars data and naming new object

#Option 1
df <- cbind(df_car_speed,car_dist)  #similarly rows could be added using rbind()
class(df)
df

#Or Option 2 - transform into a data frame before binding
df_car_dist <- as.data.frame(car_dist)  #op2
df_opt2 <- cbind(df_car_speed,df_car_dist)
class(df_opt2)
df_opt2

```

* cbind will only work if both objects have the same number of rows
* rbind could be used if you want to add a few new observations to an existing dataset (bc downloading big datasets could take time)
* good practice to turn objects into dataframe before cbind because if it is a matrix (as opposed to a vector) it might not merge correctly

Note that when we transformed the vector in a data frame the name of the vector became the column name.

```{r}
colnames(df)  #or simply names()
names(df)
```

## Creating a data frame

How would you create a data frame?

* the following function allows you to name the column

```{r}
#useful function data.frame()
create_df <- data.frame("speed"=car_speed,"dist"=car_dist)
create_df
```


## Data frame functions

Some useful functions to use with data frames.

```{r}
ncol(df)  #number of columns
nrow(df)  #number of rows
length(df)  #same as ncol - might seem like number of rows but in df - it is number of columns
summary(df) #summary statistics of columns
head(df) #show the first 6 rows of df
# sometimes you might want to look at the tail of the data (in case the last row of the data is some sort of data summary)

# If you know the number of the column you want you can refer to that to access column
# before the comma refers to rows, after the comma refers to columns

df[,1]

# if you want a certain range of rows or columns, you can use a colon
df_2 <- df[1:10,1:2]

#you could also use this notation to delete columns
df_deleted <- df[,-2]
df_deleted
```