---
title: "4: Part 1 - Data Wrangling"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2021"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Describe the usefulness of data wrangling and its place in the data pipeline
2. Wrangle datasets with dplyr functions
3. Apply data wrangling skills to a real-world example dataset

## Set up your session

Today we will work with a dataset from the [North Temperate Lakes Long-Term Ecological Research Station](https://lter.limnology.wisc.edu/about/overview). The NTL-LTER is located in the boreal zone in northern Wisconsin, USA. We will use the [chemical and physical limnology dataset](https://lter.limnology.wisc.edu/content/cascade-project-north-temperate-lakes-lter-core-data-physical-and-chemical-limnology-1984), running from 1984-2016. 

Opening discussion: why might we be interested in long-term observations of temperature, oxygen, and light in lakes?

> Add notes here: 

```{r, message = FALSE}
#getwd()
#install.packages(tidyverse)
library(tidyverse)
#install.packages(lubridate)
library(lubridate)
# stringAsFactors makes summary works for character entries
NTL.phys.data <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
# check to see data was imported successfully
colnames(NTL.phys.data)
# same date is repeated because data is recorded at different depths so there are multiple entries on each day
head(NTL.phys.data)
# head - see first few observations
summary(NTL.phys.data)
str(NTL.phys.data)
dim(NTL.phys.data)

class(NTL.phys.data$sampledate)
# Format sampledate as date
NTL.phys.data$sampledate <- as.Date(NTL.phys.data$sampledate, format = "%m/%d/%y")
```

## Data Wrangling

Data wrangling extends data exploration: it allows you to process data in ways that are useful for you. An important part of data wrangling is creating *tidy datasets*, with the following rules: 

1. Each variable has its own column
2. Each observation has its own row
3. Each value has its own cell

What is the best way to wrangle data? There are multiple ways to arrive at a specific outcome in R, and we will illustrate some of those approaches. Your goal should be to write the simplest code that will get you to your desired outcome. However, there is sometimes a trade-off of the opportunity cost to learn a new formulation of code and the time it takes to write complex code that you already know. Remember that the best code is one that is easy to understand for yourself and your collaborators. Remember to comment your code, use informative names for variables and functions, and use reproducible methods to arrive at your output.

## Dplyr Wrangling Functions

`dplyr` is a package in R that includes functions for data manipulation (i.e., data wrangling or data munging). `dplyr` is included in the tidyverse package, so you should already have it installed on your machine. The functions act as verbs for data wrangling processes. For more information, run this line of code:

```{r, results = "hide"}
vignette("dplyr")
```

### Filter

Filtering allows us to choose certain rows (observations) in our dataset.

Here are the relevant commands used in the `filter` function. Add some notes to designate what these commands mean. 
`==` - filter and compare value 
`!=` - to check if vales are different
`<` - less than
`<=` - less than or equal
`>` - 
`>=`
`&` - if you have multiple conditions to check
`|` - if one or the other should be true

```{r}
class(NTL.phys.data$lakeid)
# can only compare things that are the same class
class(NTL.phys.data$depth)
# can compare logically with numeric

# matrix filtering
# filtering only the rows that have a zero on column depth - will appear in environment
NTL.phys.data.surface1 <- NTL.phys.data[NTL.phys.data$depth == 0,]

# dplyr filtering - using the filter function
# Filter(data set, whatever you are trying to filter)
NTL.phys.data.surface2 <- filter(NTL.phys.data, depth == 0)
NTL.phys.data.surface3 <- filter(NTL.phys.data, depth < 0.25)

# Did the methods arrive at the same result? - YES!
head(NTL.phys.data.surface1)
dim(NTL.phys.data.surface1)
head(NTL.phys.data.surface2)
dim(NTL.phys.data.surface2)
head(NTL.phys.data.surface3)
dim(NTL.phys.data.surface3)

# Choose multiple conditions to filter

# summary shows us the most occurrances for the selected variable
summary(NTL.phys.data$lakename)

# how to select rows for *character* variable
# selecting paul lake and peter lake - using "or" operator which is "|"
NTL.phys.data.PeterPaul1 <- filter(NTL.phys.data, lakename == "Paul Lake" | lakename == "Peter Lake")

# another way to do this is to exclude the other lakes we are not interested in using the "!=" command
# remember that filter only chooses rows
NTL.phys.data.PeterPaul2 <- filter(NTL.phys.data, lakename != "Central Long Lake" & 
                                     lakename != "Crampton Lake" & lakename != "East Long Lake" &
                                     lakename != "Hummingbird Lake" & lakename != "Tuesday Lake" &
                                     lakename != "Ward Lake" & lakename != "West Long Lake")

# This method will be useful when you have multiple selections - just a cleaner way
NTL.phys.data.PeterPaul3 <- filter(NTL.phys.data, lakename %in% c("Paul Lake", "Peter Lake"))

# Choose a range of conditions of a numeric or integer variable
# day number will be 1-365 and will be the same year to year
summary(NTL.phys.data$daynum)
# first day of june is 152 and the first day of november is 305 so we can select this range - this is the day number equivalent of the month range we are looking for
NTL.phys.data.JunethruOctober1 <- filter(NTL.phys.data, daynum > 151 & daynum < 305)
NTL.phys.data.JunethruOctober2 <- filter(NTL.phys.data, daynum > 151, daynum < 305)
NTL.phys.data.JunethruOctober3 <- filter(NTL.phys.data, daynum >= 152 & daynum <= 304)
# another way to do this
NTL.phys.data.JunethruOctober4 <- filter(NTL.phys.data, daynum %in% c(152:304))

# Exercise: 
# filter NTL.phys.data for the year 1999
# what code do you need to use, based on the class of the variable?
class(NTL.phys.data$year4)

# Exercise: 
# filter NTL.phys.data for Tuesday Lake from 1990 through 1999.


```
Question: Why don't we filter using row numbers?

> Answer: 

### Arrange

Arranging allows us to change the order of rows in our dataset. By default, the arrange function will arrange rows in ascending order.

```{r}
NTL.phys.data.depth.ascending <- arrange(NTL.phys.data, depth)
# doesnt remove any observations - just arranges in order
NTL.phys.data.depth.descending <- arrange(NTL.phys.data, desc(depth))

# Exercise: 
# Arrange NTL.phys.data by temperature, in descending order. 
# Which dates, lakes, and depths have the highest temperatures?


```
### Select

Selecting allows us to choose certain columns (variables) in our dataset.

```{r}
# creating a new data frame, select original data, then range of columns
NTL.phys.data.temps <- select(NTL.phys.data, lakename, sampledate:temperature_C)
# same number of observations but only certain subset of data

```
### Mutate

Mutating allows us to add new columns that are functions of existing columns. Operations include addition, subtraction, multiplication, division, log, and other functions.

```{r}
# adding new column that represents farenheit
NTL.phys.data.temps <- mutate(NTL.phys.data.temps, temperature_F = (temperature_C*9/5) + 32)

```

## Lubridate

A package that makes coercing date much easier is `lubridate`. A guide to the package can be found at https://lubridate.tidyverse.org/. The cheat sheet within that web page is excellent too. This package can do many things (hint: look into this package if you are having unique date-type issues), but today we will be using two of its functions for our NTL dataset. 

```{r}
# add a month column to the dataset
# first - create new column that extracts just the month information from column sample date which is already a date object
NTL.phys.data.PeterPaul1 <- mutate(NTL.phys.data.PeterPaul1, month = month(sampledate)) 

# reorder columns to put month with the rest of the date variables
# we want month close to columns that have date
# select (date set, what you want first, then month, then the rest)
NTL.phys.data.PeterPaul1 <- select(NTL.phys.data.PeterPaul1, lakeid:daynum, month, sampledate:comments)

# find out the start and end dates of the dataset
# pull first row and last row
interval(NTL.phys.data.PeterPaul1$sampledate[1], NTL.phys.data.PeterPaul1$sampledate[21613])
# if your date is not ordered by date, you can find the interval by using function first and last (part of dplyr package)
# this will find the first and last date - much better way because it will work for any size and any order
interval(first(NTL.phys.data.PeterPaul1$sampledate), last(NTL.phys.data.PeterPaul1$sampledate))
```


## Pipes

Sometimes we will want to perform multiple functions on a single dataset on our way to creating a processed dataset. We could do this in a series of subsequent functions or create a custom function. However, there is another method to do this that looks cleaner and is easier to read. This method is called a pipe. We designate a pipe with `%>%`. A good way to think about the function of a pipe is with the word "then." 

Let's say we want to take our raw dataset (NTL.phys.data), *then* filter the data for Peter and Paul lakes, *then* select temperature and observation information, and *then* add a column for temperature in Fahrenheit: 

```{r}
# because we are doing this in a pipe, we dont have to specify data frame name when we pull columns
NTL.phys.data.processed <- 
  NTL.phys.data %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") %>%
  select(lakename, sampledate:temperature_C) %>%
  mutate(temperature_F = (temperature_C*9/5) + 32)
  
```

Notice that we did not place the dataset name inside the wrangling function but rather at the beginning.

### Saving processed datasets

```{r}
write.csv(NTL.phys.data.PeterPaul1, row.names = FALSE, file = "./Data/Processed/NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv")
```

## Closing Discussion

When we wrangle a raw dataset into a processed dataset, we create a code file that contains only the wrangling code. We then save the processed dataset as a new spreadsheet and then create a separate code file to analyze and visualize the dataset. Why do we keep the wrangling code separate from the analysis code?


