---
title: "5: Lab - Data Visualization Basics"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Answer questions on M4/A4
2. Answer questions on M5 videos
3. Additional comments on videos - expression() and saving plot
4. Perform simple data visualizations in the R package `ggplot`

## Expression function and Saving plots

The expression function can be used when you want to add mathematical symbols and subscripts mixed with regular letter to labels. 

This shows how you add an expression to a plot and specify where on the plot you want to add the expression.

This also shows how you would save the plot to the output folder.

```{r}
library(ggplot2)
Plot_text <-
ggplot(data=data.frame(x=0,y=0))+
  geom_point(aes(x=x,y=y)) +
  geom_text(x=0,y=0.025,label=expression(paste("Mean is ", mu,",", R^{2},'=0.6')))
print(Plot_text)

ggsave("./Output/Plot_text.jpg",Plot_text, height = 4, width = 6, units = "in", dpi = 300)

```

## Import Datasets

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)
library(ggridges)
library(viridis)
library(RColorBrewer)
library(colormap)

PeterPaul.chem.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv", stringsAsFactors = TRUE)

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")

```


## Plotting continuous variables over time: Scatterplot and Line Plot

Exercise: build your own scatterplots of PeterPaul.chem.nutrients

```{r, fig.height = 3, fig.width = 4, warning = FALSE}

# 1. 
# Plot surface temperatures by day of  year. 
TempvDay <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = daynum, y = temperature_C)) +
  geom_point()
print(TempvDay)

# Color your points by year, and facet by lake in two rows.
TempvDayvLake <-
  ggplot(PeterPaul.chem.nutrients, aes(x = daynum, y = temperature_C, color = year4, shape = lakename)) +
  geom_point() 
print(TempvDayvLake)

TDplot.faceted <-
  ggplot(PeterPaul.chem.nutrients, aes(x = daynum, y = temperature_C, color = year4)) +
  geom_point() +
  facet_wrap(vars(lakename), nrow = 2)
print(TDplot.faceted)

# Change the ylab name

TDplot.faceted.2 <-
  ggplot(PeterPaul.chem.nutrients, aes(x = daynum, y = temperature_C, color = year4)) +
  geom_point() +
  facet_wrap(vars(lakename), nrow = 2) +
  ylab("Temperature")
print(TDplot.faceted.2)

# to plot only surface temperature - filter where depths = 0
# could also filter before pulling it in
# if you want to color by factor, the color function has to be in aes()
# you can also add the color function in geom_point()
TDplot.faceted.3 <-
  ggplot(subset(PeterPaul.chem.nutrients, depth == 0), aes(x = daynum, y = temperature_C, color = year4)) +
  geom_point() +
  facet_wrap(vars(lakename), nrow = 2) +
  ylab("Temperature")
print(TDplot.faceted.3)

#2. 
# Plot temperature by date. Color your points by depth.
TempvDate <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = sampledate, y = temperature_C)) +
  geom_point(aes(color = depth)) +
  ylab("Temperature in Celsius")
print(TempvDate)

# Change the size of your point to 0.5
TempvDate.1 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = sampledate, y = temperature_C)) +
  geom_point(aes(color = depth), size = 0.5) +
  ylab("Temperature in Celsius")
print(TempvDate.1)

# Change the color palette to magma and play with direction (+- 1), which one makes more sense?
  # makes more sense to have minus one because the colors represent the depth more - GOOD TIPS FOR       VISUALIZATOIN
TempvDate.2 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = sampledate, y = temperature_C)) +
  geom_point(aes(color = depth), size = 0.5) +
  ylab("Temperature in Celsius") +
  scale_color_viridis(option = "magma", direction = 1)
print(TempvDate.2)

# Change x axis to include marker/labels every 5 years
TempvDate.3 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = sampledate, y = temperature_C)) +
  geom_point(aes(color = depth), size = 0.5) +
  ylab("Temperature in Celsius") +
  scale_color_viridis(option = "magma", direction = -1) +
  scale_x_date(
    date_breaks = "5 years", date_labels = "%Y"
  )
print(TempvDate.3)



```
## Plotting the relationship between two continuous variables: Scatterplot

Exercise: Plot relationships between air quality measurements

```{r, fig.height = 3, fig.width = 4, warning = FALSE}

# 3. 
# Plot AQI values for ozone by PM2.5, colored by latitude 
AQIvPM2.5 <- 
  ggplot(EPAair, aes(x = PM2.5, y = Ozone)) +
  geom_point(aes(color = meanLat))
print(AQIvPM2.5)

# Make the points 50 % transparent
AQIvPM2.5.1 <- 
  ggplot(EPAair, aes(x = PM2.5, y = Ozone)) +
  geom_point(aes(color = meanLat), alpha = 0.5)
print(AQIvPM2.5.1)

# Add a line of best fit for the linear regression of these variables.
# can define the color of the best fit line
# have to use expression function in ylab if you are creating special characters
AQIvPM2.5.2 <- 
  ggplot(EPAair, aes(x = PM2.5, y = Ozone)) +
  geom_point(aes(color = meanLat), alpha = 0.5)+
  geom_smooth(method = lm, color = "red") +
  ylab(expression("O"[3]* " AQI Value"))
print(AQIvPM2.5.2)



```

## Plotting continuous vs. categorical variables

Exercise: Plot distributions of AQI values for EPAair.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}

# 4. 
# Create several types of plots depicting PM2.5, divided by year. 
# Choose which plot displays the data best and justify your choice. 
PMPlot.faceted <- 
  ggplot(EPAair, aes(x = Month, y = PM2.5)) +
  geom_point() +
  facet_grid(Site.Name ~ Year) +
  scale_x_continuous(limits = c(1,12), breaks = c(1:12))
print(PMPlot.faceted)


```


## Visualization challenge

The following graph displays the counts of specific endpoints measured in neonicotinoid ecotoxicology studies. The way it is visualized, however, is not effective. Make the following coding changes to improve the graph: 

1. Change the ordering of the "Endpoint" factor (function: `reorder`) so that the highest counts are listed first (hint: FUN = length)
2. Plot the barplot with the reordered factor levels. Add this line of code to make the bars show up left to right:   scale_x_discrete(limits = rev(levels(Neonics$Endpoint)))
3. Adjust the x axis labels so they appear at a 45 degree angle. 
4. Change the color and/or border on the bars. Should you have a consistent color across all bars, or a different color for each bar?
```{r}

Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv")
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint))



```
